Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-01 paper/zhang-cook-laa-langrene-menendez.tex
read through everything before section vis diag

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While many indices
have been proposed in the literature, less work has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object for the optimisation in the guided
tour and introduce visual diagnostics based on the data object
collected. These diagnostics and workflows can be applied to a broad
class of optimisers, to assess their performance. An R package,
\pkg{ferrn}, has been created to implement the diagnostics.
}

@@

Visualisation is widely used in exploratory data analysis
\citep{tukey1977exploratory, unwin2015graphical, healy2018data, wilke2019fundamentals}.
Presenting information in graphics often unveils information that would
otherwise not be aware of and provides a more comprehensive
understanding of the problem at hand. Task specific tools presented by
\citet{li2020visualizing} show how visualisation can be used to
understand the behaviour of neural network on classification models, but
no general visualisation tool available for diagnosing optimisation
procedure. The work presented in this paper brings visualization tools
into optimisation problems with an aim to better understand the
performance of the optimisers in practice.

@@

The focus of this paper is on the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational} which is an
exploratory data analysis tool that is defined to detect
\emph{interesting structures} or features in high-dimensional data
through a set of lower-dimensional projections that cover the entire
high dimensional space using interpolation methods called tours
\citep{cook2008grand}. The target of the optimisation is to identify the
most \emph{interesting} low-dimensional views of the data given by a
corresponding projection matrix. The most \emph{interesting} structures
are formally defined by a function of projections, called index function
which is optimized to uncover the most revelling structures in a high
dimensional space \citep{cook1993projection}.

The optimization challenges encountered in the projection pursuit guided
tour problem are common to those of optimization in general. Examples of
those include the existence of multiple maxima (local and global), the
trade off between computational burden and proximity to the maxima,
dealing with noisy objective functions that might be non smooth and non
differentiable \citep{jones1998efficient}. Those are not unique to this
context and therefore the visualization tools and optimization methods
presented in this paper can be easily applied to any other optimization
problems.

The remainder of the paper is organised as follows. Section \ref{optim}
provides an overview of optimisation methods, specifically line search
methods. Section \ref{tour} reviews projection pursuit guided tour,
defines the optimisation problem and introduces three existing
algorithms. Section \ref{vis-diag} presents the new visual diagnostics.
A data structure is defined to capture information during the
optimisation, and used in different types of diagnostic plots. Section
\ref{application} shows applications of how these plots can be used to
understand and compare different algorithms. We also discuss how these
insights contribute to modifications that improve the algorithms.
Finally, Section \ref{implementation} describes the R package:
\pkg{ferrn}, that implements the visual diagnostics.

@@

The projection pursuit guided tour combines two different methods in
exploratory data analysis, focusing on different aspects. Projection
pursuit, coined by \citet{friedman1974projection}, detects interesting
structures (e.g.~clustering, outliers and skewness) in multivariate data
via low dimensions projection. The guided tour is using ideas from
projection pursuit to define a particular variation in a broader class
of data visualisation methods, building on the grand tour approach
\citep{As85}.

To define projection pursuit, we first need to establish the notation
used. Let \(\mathbf{X}_{n \times p}\) be the data matrix, with \(n\)
observations in \(p\) dimensions. A d-dimensional projection can be seen
as a linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\),
and defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection matrix. Define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) (corresponding to
an associated projection matrix \(\mathbf{A}\)) onto an index value
\(I\) (QUESTION: Isn't f the index function?). This is commonly known as
the projection pursuit index function, or just index function, and is
used to measure the ``interestingness'' of a given projection.

A number of index functions have been proposed in the literature to
detect different data structures, including Legendre index
\citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, natural Hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} for
supervised classification problems, and PDA index
\citep{lee2010projection}, which is an extension of the LDA index.

As a general visualisation method, a tour produces animations of high
dimensional data via rotations between low dimension planes. Different
tour types choose these planes differently, for example, a grand tour
\citep{cook2008grand} selects the planes randomly to provide a general
overview and a manual tour \citep{cook1997manual} gradually phases in
and out one variable, to understand the contribution of that variable in
the projection. Guided tour, the main interest of this paper, chooses
planes with the aid of projection pursuit to gradually reveal the most
interesting projection in the low dimension space. Given a random start,
projection pursuit iteratively finds bases with higher index values and
the guided tour constructs the geodesic interpolation between these
planes to form a tour path. Intuitively, Figure \ref{fig:tour-path}
shows a sketch of the tour path where the blue frames are produced by
the projection pursuit optimisation algorithm, and the white frames
interpolate between them. Mathematical details of the geodesic
interpolation can be found in \citet{buja2005computational}. The tour
method has been implemented in the R package \CRANpkg{tourr}
\citep{tourr}.

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.5\linewidth,height=0.25\textheight]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/img/tour_path} \caption[Each square (frame) represents the projected data with a corresponding basis]{Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.}\label{fig:tour-path}
\end{figure}
\end{Schunk}

@@

The optimisation problem in the tour context is stated as follows: Given
a randomly generated starting basis \(\mathbf{A}_1\), projection pursuit
finds the final projection basis \(\mathbf{A}_T\) that satisfies the
following optimisation problem:

@@

\noindent where \(I_d\) is the \(d\)-dimensional identity matrix and the
constraint requires the projection bases \(\mathbf{A}\) to be orthogonal
matrices.

Several features of this optimisation are worth noticing. First of all,
this is a constrained optimisation problem as the decision variables
form the entries of a projection basis, which is required to be
orthonormal. It is also likely that the objective function may not be
differentiable for a constructed index function and in these cases,
gradient-based methods may not work well. Although finding the global
maximum is the goal of an optimisation problem, it is also interesting
to inspect local maximum in projection pursuit since it could present
unexpected interesting projections. Lastly, there is also one
computational consideration: the optimisation procedure needs to be fast
to compute since the tour animation is played in real-time.

@@

Below we introduce three line search algorithms in tour: simulated
annealing (SA), simulated annealing with jump out (SAJO), and pseudo
derivative (PD).

@@

Simulated annealing (SA) is a random search device that samples a
candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the current
basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is a randomly generated matrix with the
same dimension as \(\mathbf{A}_{\text{cur}}\). \(\mathbf{A}_{l}\) is
then orthogonalised to ensure the orthonormal constraint is fulfilled.
When a basis is found with index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the search terminates and outputs the basis
for guided tour to construct an interpolation path. The next iteration
of search begins after adjusting \(\alpha\) by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\). The termination condition
is when the maximum number of iteration \(l_{\max}\) is reached. The
algorithm of simulated annealing is summarised in Algorithm
\ref{random-search}. A slightly different cooling scheme has been
proposed by \citet{posse1995projection} to avoid the search space being
reduced too fast. A halving parameter \(c\) is introduced and \(\alpha\)
is only adjusted if the last search takes more than \(c\) times to find
an accepted basis.

@@

Simulated annealing with jump out (SAJO)
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process but allows a probabilistic acceptance of a
inferior basis with lower index value based on the annealing \(T(l)\).
Given an initial \(T_0\), the temperature at iteration \(l\) is defined
as \(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to
have an index value larger than the current basis, simulated annealing
gives it a second chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the algorithm to jump out of a local maximum and
enables a more holistic search of the whole parameter space. This
feature is particularly useful when local maxima are present. The
algorithm \ref{simulated-annealing} highlights how SAJO differs from SA
in the inner loop.

@@

In pseudo derivative search \citep{cook1995grand}, the search direction
is computed using the most prominent direction that deviating an tiny
angle of \(\delta\) from the current basis. The step size is chosen by
optimising the index value along the geodesic direction over an 90
degree angle from \(-\pi/4\) to \(\pi/4\) along the search direction
chosen. The optima \(\mathbf{A}_{**}\) is returned for the current
iteration if it meets the percentage improve condition or when
\(l_{\max}\) is reached. Algorithm \ref{search-geodesic} summarises the
inner loop of the pseudo derivative search.

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-search-1} \caption[A comparison of search by two otimisers]{A comparison of search by two otimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iteration because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-interp-1} \caption[Trace plots of the interpolating bases with the same optimisation routine as the previous figure]{Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotinic.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-1} \caption[The PCA plot of 1D projection problem on the 5-variable dataset, \code{boa5}, using the holes indexwith two optimisers]{The PCA plot of 1D projection problem on the 5-variable dataset, \code{boa5}, using the holes indexwith two optimisers: SA and PD. All the bases in PD has been flipped and a grey dashed line has been annotated to indicate the symmetry of two starting bases.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-animated-1} \caption{Six selected frames from the animated PCA plots in the previous figure. With animation, the start and end position of each optimisation is easier to identify.  A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504242845/b73f37175a}}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-tour-1} \caption{Six selected frames from the tour animation for viewing the same two optmisations as previous two figures. The tour animation allows for an appreciation of the search bases in the original high dimensional space. A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504328122/9be84db563}}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/interruption-1} \caption[Trace plot of 2D projection on \code{boa6} data with holes index, optimised by optimiser SA]{Trace plot of 2D projection on \code{boa6} data with holes index, optimised by optimiser SA.}\label{fig:interruption}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/polish-1} \caption[Trace plot of 2D projection on \code{boa6} data with holes index, optimised first by SA and then by polish search]{Trace plot of 2D projection on \code{boa6} data with holes index, optimised first by SA and then by polish search. The polish step results in clearer cut on the edges of the four clusters in the data structure.}\label{fig:polish}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/flip-sign-1} \caption[PCA plot of 1D projection of boa6 data, optimised by SA, with seed 2463]{PCA plot of 1D projection of boa6 data, optimised by SA, with seed 2463.}\label{fig:flip-sign}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/noisy-better-geo-1} \caption[Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimiers]{Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimiers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO success in optimise the index, SAJO takes much longer than SA and finishes off closer to the theoretial best.}\label{fig:noisy-better-geo}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-result-1} \caption[PCA plot of 1D projection on the 6\-variable data, \code{boa6}, with normal kolmogorov index]{PCA plot of 1D projection on the 6\-variable data, \code{boa6}, with normal kolmogorov index. Two optimisers: SA and SAJO and two search sizes: 0.5 and 0.7 are used in the simulation for 20 replicates. Local optimum (x), is presented in this experiment, along with the global optimum (*). The final basis is larger and more opaque than the start basis for each path and color represents whether the global or local optimum is found, after polishing.}\label{fig:kol-result}
\end{figure}
\end{Schunk}

@@

  \begin{itemize}
  \tightlist
  \item
    \code{explore\_trace\_search()} produces summary plots, as shown in
    Figure \ref{fig:toy-search}
  \item
    \code{explore\_trace\_interp()}produces trace plots for the
    interpolation points, as shown in Figure \ref{fig:toy-interp}
  \item
    \code{explore\_space\_pca()} produces plots of projection basis on
    the reduced space by PCA, as shown in Figure \ref{fig:toy-pca}.
    Animated version in Figure \ref{fig:toy-pca-animated} can be turned
    on via the argument \texttt{animate\ =\ TRUE}
  \item
    \code{explore\_space\_tour()} produces animated tour view on the
    full space of the projection bases, as shown in Figure
    \ref{fig:toy-tour}.
  \end{itemize}
\item
  \code{get\_*()} extracts and manipulates certain components from the
  existing data object.

  \begin{itemize}
  \tightlist
  \item
    \code{get\_best()} extracts the best basis found in the data object
  \item
    \code{get\_start()} extracts the starting basis
  \item
    \code{get\_interp()} extracts the observations in the interpolation
  \item
    \code{get\_interp\_last()} extracts the end observations of the
    interpolation in each iteration
  \item
    \code{get\_anchor()} extracts the target observations found by the
    optimiser
  \item
    \code{get\_search()} extracts the search observations evaluated by
    the optimiser
  \item
    \code{get\_search\_count()} produces the summary table of the number
    of observation in each iteration
  \item
    \code{get\_centre()} produces the centre point of the basis space
    estimated by the starting points
  \item
    \code{get\_space\_param()} produces the coordinates of the centre
    and radius of the basis space
  \item
    \code{get\_theo()} extracts the theoretical observations from the
    data object
  \item
    \code{get\_interrupt()} extract the end point of the interpolation
    and the target point when an interruption happens
  \item
    \code{get\_basis\_matrix()}: flattens all the bases into a matrix
  \end{itemize}
\item
  \code{bind\_*()} incorporates additional information outside the tour
  optimisation into the data object.

  \begin{itemize}
  \tightlist
  \item
    \code{bind\_theoretical()} incorporates the best possible basis to
    the existing data object with the supply of the index function and
    original data for producing the index value.
  \item
    \code{bind\_random()} generates 1000 points on the high dimensional
    surface of a sphere and binds it to the existing data object and
    output as a tibble object.
  \item
    \code{bind\_random\_matrix()} binds the points to the basis matrix.
  \end{itemize}
\item
  Utilities

  \begin{itemize}
  \tightlist
  \item
    \code{add\_*()} are internal wrapper functions that facilitate the
    composition of PCA plot
  \item
    \code{theme\_fern()} and \code{format\_label()} for better display
    of the grid lines and axis formatting
  \item
    \code{clean\_method()} for clean up the name of the optimisers
  \item
    \code{botanical\_palettes} is a collection of colour palettes from
    Australian native plants. Quantitative palettes include daisy,
    banksia and cherry and sequential palettes contain fern and acacia.
  \item
    \code{botanical\_pal()} as the colour interpolator
  \end{itemize}
\end{itemize}

>>>>

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projeciton
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and workflows can be applied to a
broad class of optimisers, to assess their performance. An R package,
\pkg{ferrn}, has been created to implement the diagnostics.
}

@@

Visualisation is widely used in exploratory data analysis
\citep{tukey1977exploratory, unwin2015graphical, healy2018data, wilke2019fundamentals}.
Presenting information in graphics often unveils information that would
otherwise not be aware of and provides a more comprehensive
understanding of the problem at hand. Task specific tools such as
\citet{li2020visualizing} show how visualisation can be used to
understand the behaviour of neural network on classification models, but
no general visualisation tool is available for diagnosing optimisation
procedures. The work presented in this paper brings visualization tools
into optimisation problems with an aim to better understand the
performance of the optimisers in practice.

@@

The focus of this paper is on the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational}, an
exploratory data analysis tool used for detecting interesting structures
in high-dimensional data through a set of lower-dimensional projections
\citep{cook2008grand}. The goal of the optimisation is to identify the
projection, characterised by the projection basis, that gives the most
interesting low-dimensional view. The interestingness of the structure
is defined by the index function, a function of the projection basis.

The optimization challenges encountered in the projection pursuit guided
tour problem are common to those of optimization in general. Examples of
those include the existence of multiple optima (local and global), the
trade off between computational burden and proximity to the optima,
dealing with noisy objective functions that might be non-smooth and
non-differentiable \citep{jones1998efficient}. Those are not unique to
this context and therefore the visualization tools and optimization
methods presented in this paper can be easily applied to any other
optimization problems.

The remainder of the paper is organised as follows. Section \ref{optim}
provides an overview of optimisation methods, specifically line search
methods. Section \ref{tour} reviews projection pursuit guided tour,
defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data
structure is defined to capture information during the optimisation, and
different diagnostic plots are designed with the data collected. Section
\ref{application} shows applications of how these diagnostic plots can
be used to discover interesting aspects of different optimisers and
guide improvement to the existing algorithms. Finally, Section
\ref{implementation} describes the R package: \pkg{ferrn}, that
implements the visual diagnostics.

@@

Projection pursuit guided tour combines two different methods
(projection pursuit and guided tour) in exploratory data analysis.
Projection pursuit, coined by \citet{friedman1974projection}, detects
interesting structures (e.g.~clustering, outliers and skewness) in
multivariate data via low dimensions projection. Guided tour is one
variation of a broader class of data visualisation methods, tour, which
visualises high-dimensional data through a series of animated
projection.

Notation is first established before defining projection pursuit. Let
\(\mathbf{X}_{n \times p}\) be the data matrix with \(n\) observations
in \(p\) dimensions. A d-dimensional projection is a linear
transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\), and defined
as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis Define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto an scalar.
This is commonly known as the projection pursuit index function, or just
index function, and is used to measure the ``interestingness'' of a
given projection. A number of index functions have been proposed in the
literature to detect different data structures
\citep[friedman1974projection;][]{hall1989polynomial, cook1993projection, posse1995projection}
and to classify groups \citep{lee2005projection, lee2010projection}.

As a general visualisation method, tour produces animations of high
dimensional data via rotations low dimension planes. Different tour
types choose these planes differently: grand tour \citep{cook2008grand}
selects the planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs the
geodesic interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The tour method has been implemented in
the R package \CRANpkg{tourr} \citep{tourr}.

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=0.5\linewidth,height=0.25\textheight]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/img/tour_path}

}

\caption[Each square (frame) represents the projected data with a corresponding basis]{Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.}\label{fig:tour-path}
\end{figure}
\end{Schunk}

@@

The optimisation problem in projection pursuit is stated as follows:
Given a randomly start basis \(\mathbf{A}_1\), projection pursuit finds
the final projection basis \(\mathbf{A}_T\) that satisfies the following
optimisation problem:

@@

\noindent where \(I_d\) is the \(d\)-dimensional identity matrix and the
constraint requires the projection bases \(\mathbf{A}\) to be
orhtonormal.

Several features of this optimisation are worth-noticing. 1) The
optimisation is constrained and the space of the bases is a
d-dimensional sphere. 2) The objective function may not be
differentiable for a constructed index function. 3) Although finding the
global maximum is the goal of an optimisation problem, it is also
interesting to inspect local maximum in projection pursuit since it
could present unexpected interesting projections. 4) The optimisation
procedure needs to be easy to compute since the tour animation needs to
be played in real-time during the optimisation.

@@

Three optimisers have been implemented in the \CRANpkg{tourr}
\citep{tourr} package: simulated annealing (SA), simulated annealing
with jump out (SAJO), and pseudo derivative (PD). Simulated annealing
(SA) is a random search optimiser that samples a candidate basis
\(\mathbf{A}_{l}\) in the neighbourhood of the current basis
\(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is generated randomly. \(\mathbf{A}_{l}\)
is then orthonormalised to fulfil the basis constraint. If
\(\mathbf{A}_{l}\) has an index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the optimiser outputs \(\mathbf{A}_{l}\)
for guided tour to construct an interpolation path. The neighbourhood
parameter \(\alpha\) is adjusted by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\) before the next iteration
starts. The optimiser terminates when the maximum number of iteration
\(l_{\max}\) is reached before a better basis can be found. The
algorithm of SA is summarised in Algorithm \ref{random-search}.
\citet{posse1995projection} has proposed a slightly different cooling
scheme by introducing a halving parameter \(c\). In his proposal, the
\(\alpha\) is only adjusted if the last iteration takes more than \(c\)
times to find a better basis.

@@

Simulated annealing with jump out (SAJO)
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process as SA but allows a probabilistic acceptance of a
basis with lower index value than the current. Given an initial value of
\(T_0\), the temperature at iteration \(l\) is defined as
\(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have
an index value larger than the current basis, SAJO gives it a second
chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the optimiser to make a move and explore the basis
space even the landing basis does have a higher index value and hence
enable the optimiser to jump out of a local optimum. The algorithm
\ref{simulated-annealing} highlights how SAJO differs from SA in the
inner loop.

@@

Pseudo derivative (PD) search \citep{cook1995grand} uses a different
strategy than SA and SAJO. Rather than randomly sample the basis space,
PD first computes a search direction by evaluating vases close to the
current basis. A step size is then chosen along the geodesic direction
by another optimisation over an 90 degree angle from \(-\pi/4\) to
\(\pi/4\). The resulting candidate basis \(\mathbf{A}_{**}\) is returned
for the current iteration if it has a higher index value than the
current one. Algorithm \ref{search-geodesic} summarises the inner loop
of the PD.

@@

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-search-1}

}

\caption[A comparison of search by two otimisers]{A comparison of search by two otimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iteration because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-interp-1}

}

\caption[Trace plots of the interpolating bases with the same optimisation routine as the previous figure]{Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotinic.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-1}

}

\caption[The PCA plot of 1D projection problem on the 5-variable dataset, \code{boa5}, using the holes indexwith two optimisers]{The PCA plot of 1D projection problem on the 5-variable dataset, \code{boa5}, using the holes indexwith two optimisers: SA and PD. All the bases in PD has been flipped and a grey dashed line has been annotated to indicate the symmetry of two starting bases.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-animated-1}

}

\caption{Six selected frames from the animated PCA plots in the previous figure. With animation, the start and end position of each optimisation is easier to identify.  A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504242845/b73f37175a}}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-tour-1}

}

\caption{Six selected frames from the tour animation for viewing the same two optmisations as previous two figures. The tour animation allows for an appreciation of the search bases in the original high dimensional space. A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504328122/9be84db563}}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/interruption-1}

}

\caption[Trace plot of 2D projection on \code{boa6} data with holes index, optimised by optimiser SA]{Trace plot of 2D projection on \code{boa6} data with holes index, optimised by optimiser SA.}\label{fig:interruption}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/polish-1}

}

\caption[Trace plot of 2D projection on \code{boa6} data with holes index, optimised first by SA and then by polish search]{Trace plot of 2D projection on \code{boa6} data with holes index, optimised first by SA and then by polish search. The polish step results in clearer cut on the edges of the four clusters in the data structure.}\label{fig:polish}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/flip-sign-1}

}

\caption[PCA plot of 1D projection of boa6 data, optimised by SA, with seed 2463]{PCA plot of 1D projection of boa6 data, optimised by SA, with seed 2463.}\label{fig:flip-sign}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/noisy-better-geo-1}

}

\caption[Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimiers]{Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimiers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO success in optimise the index, SAJO takes much longer than SA and finishes off closer to the theoretial best.}\label{fig:noisy-better-geo}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-result-1}

}

\caption[PCA plot of 1D projection on the 6\-variable data, \code{boa6}, with normal kolmogorov index]{PCA plot of 1D projection on the 6\-variable data, \code{boa6}, with normal kolmogorov index. Two optimisers: SA and SAJO and two search sizes: 0.5 and 0.7 are used in the simulation for 20 replicates. Local optimum (x), is presented in this experiment, along with the global optimum (*). The final basis is larger and more opaque than the start basis for each path and color represents whether the global or local optimum is found, after polishing.}\label{fig:kol-result}
\end{figure}
\end{Schunk}

@@

  \begin{itemize}
  \tightlist
  \item
    \code{explore\_trace\_search()} produces summary plots in Figure
    \ref{fig:toy-search}
  \item
    \code{explore\_trace\_interp()}produces trace plots for the
    interpolation points in Figure \ref{fig:toy-interp}
  \item
    \code{explore\_space\_pca()} produces the PCA plot of projection
    bases on the reduced space in Figure \ref{fig:toy-pca}. Animated
    version in Figure \ref{fig:toy-pca-animated} can be turned on via
    the argument \texttt{animate\ =\ TRUE}
  \item
    \code{explore\_space\_tour()} produces animated tour view on the
    full space of the projection bases in Figure \ref{fig:toy-tour}.
  \end{itemize}
\item
  \code{get\_*()} extracts and manipulates certain components from the
  existing data object.

  \begin{itemize}
  \tightlist
  \item
    \code{get\_anchor()} extracts target observations
  \item
    \code{get\_basis\_matrix()} flattens all the bases into a matrix
  \item
    \code{get\_best()} extracts the observation with the highest index
    value in the data object
  \item
    \code{get\_dir\_search()} extracts directional search observations
    for PD search
  \item
    \code{get\_interp()} extracts interpolated observations
  \item
    \code{get\_interp\_last()} extracts the ending interpolated
    observations in each iteration
  \item
    \code{get\_interrupt()} extracts the ending interpolated
    observations and the target observations if the interpolation is
    interrupted
  \item
    \code{get\_search()} extracts search observations
  \item
    \code{get\_search\_count()} extracts the count of search
    observations
  \item
    \code{get\_space\_param()} produces the coordinates of the centre
    and radius of the basis space
  \item
    \code{get\_start()} extracts the starting observation
  \item
    \code{get\_theo()} extracts the theoretical best observations, if
    presented
  \end{itemize}
\item
  \code{bind\_*()} incorporates additional information outside the tour
  optimisation into the data object.

  \begin{itemize}
  \tightlist
  \item
    \code{bind\_theoretical()} binds the theoretical best observation in
    simulated experiment
  \item
    \code{bind\_random()} binds randomly generated bases in the
    projection bases space to the data object
  \item
    \code{bind\_random\_matrix()} binds randomly generated bases and
    outputs in a matrix format
  \end{itemize}
\item
  Utilities

  \begin{itemize}
  \tightlist
  \item
    \code{add\_*()} are internal wrapper functions that facilitate the
    composition of PCA plot
  \item
    \code{theme\_fern()} and \code{format\_label()} for better display
    of the grid lines and axis formatting
  \item
    \code{clean\_method()} for clean up the name of the optimisers
  \item
    \code{botanical\_palettes} is a collection of colour palettes from
    Australian native plants Quantitative palettes include daisy,
    banksia and cherry and sequential palettes contain fern and acacia
  \item
    \code{botanical\_pal()} as the colour interpolator
  \item
    \code{scale\_color\_*()} and \code{scale\_fill\_*()} for scaling the
    colour and fill of the plot
  \end{itemize}
\end{itemize}
